Human capital mediates natural selection in contemporary humans

### 2022-03-14 response-to-reviewers-BE.Rmd

---
output: 
  pdf_document:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
mainfont: Baskerville
---

@@

```

Dear Professor Keller,

Thank you very much for the chance to revise our manuscript. We would also like to thank both reviewers for their comments. Below we deal with the points they raised.

Code to produce this paper is now publicly available at <https://github.com/hughjonesd/negative-selection>. XXX

With best wishes,

David Hugh-Jones\
Abdel Abdellaoui

@@

> Most curious is the failure to cite Fisher (1930).... The overlap in theory and observations with the present work is striking, so much so that the omission is equally striking.

We thank the reviewer for giving us this reference. Indeed it is striking. Fisher's theory about the positive correlation he mentions is that it's driven by selection bias: Yale parents who can afford to send 6 children to Yale are "more able" than those who can only send 1, and as a result, among Yale students, children from bigger families are smarter. This is related to the income effect in our theory. We now mention this work in our conclusion.

> The authors do not have to respond to my next point. This paper already has a lot. But I am curious as to whether the authors have an opinion about Fisher's proposed solution to this problem (if it is a problem). He in essence suggested that the population be stratified into classes based on "human capital," as we might say now, and that within each stratum redistribution should be performed so as to reward parents of larger families. The basic idea is that if I have four children, while an academic colleague who is the same age as me has none (which has allowed him to publish twice as many papers), the tax-and-spend system will take away some of his income and give it to me, the total amount of redistribution being such as to remove the fundamental tradeoff. Can this work in theory? Can it be made politically feasible?

Indeed, we don't want to talk about policy in the paper - we think we do not yet know enough. For what it's worth, here are my personal thoughts as corresponding author. First, government schemes to encourage fertility have a poor track record, see e.g. Gauthier (2007). Second, my strong gut feeling is that it is wrong for governments to (bluntly put) institute breeding programs for their citizens. I wrote more informally on this topic [here](https://wyclif.substack.com/p/the-natural-selection-paper-part-822?s=w).

@@

> Another general point is that there is some uncertainty over whether the fertility-IQ correlation is negative in entire populations. Some suggest that its negative value in unrepresentative samples is the result of ascertainment bias (e.g., Higgins, Reed, & Reed, 1962; Kolk & Kieron, 2019). In fact, I have reviewed a manuscript where the authors declined using the UK Biobank for a particular analysis (while using it for all others) on the grounds that the fertility-IQ relationship cannot be accurately inferred in that dataset. My position is that the fertility-IQ correlation has been negative, at most times since relevant data have been collected, while being less so at certain atypical times (e.g., the Baby Boom) and in men. In the present work, the authors use sample reweighting to address ascertainment bias in the UK Biobank, but this is not guaranteed to be effective. A brief literature review addressing the sign of the fertility-IQ correlation, in most Western populations in recent decades, therefore seems warranted to provide some reassurance. My suggestion is to begin with Lynn and Van Court (2004), Meisenberg (2010), and Reeve, Heeney, and Woodley (2018).

We agree this is an important topic, but we are more focused on human capital and earnings than IQ specifically. (Of course the polygenic score for educational attainment does also predict IQ.) The negative fertility-income relationship, in developed countries, is a very well-established stylized fact, and we mention some papers that also show income is causal. We now mention two of the cited papers.

Regarding ascertainment bias, the reviewer is right that there are no guarantees we can eliminate it. When we do correct for ascertainment bias, every correction increases our estimated effect sizes. So we expect that any remaining ascertainment is also probably biasing our estimates towards zero. We now say this.

@@

We have tweaked the discussion slightly to reflect the above. The best way to gain intuition is probably to try a few numerical examples.

## Reviewer 2

> 1\. The authors have improved the presentation of their results, but it can still be further improved. There are still so many results, plus a model, and as a result the paper still reads like a disconnected set of findings and lacks an overall coherent narrative (although a bit less so than before).

We have tried to work on this. Unconnected results and minor points have been pushed to footnotes and/or the appendix.

We don't want to go too far, by pretending that the theory came to us fully-formed in our ivory tower, and then we tested it. In fact, the motivating results led us to develop the theory, first informally and then more precisely; only the results in section 4 should be considered a clean test.

@@

We think these analyses are important. First, they motivate our theory. As we say, though the mediating variables are endogenous, we build a theory in which (e.g.) education is an endogenous choice variable.

We also think the results matter in their own right. Any theory of contemporary natural selection needs to explain why correlations with fertility are lower at higher levels of income, education, etc. Non-causal explanations would be fine, but theories which don't generate these correlations must be missing something. Note that Fisher, cited by Reviewer 1, used observed correlations quite like ours to differentiate between theories of the fertility-income relationship. Generally, significant correlations have often led to important theoretical literatures, even when the direction of causality is uncertain (e.g. education/wages in labour economics, or democracy/GDP levels in political economy).

As described above, we have shortened Section 2 to focus on the key results of variation by income, education, number of partners, partner presence and AFLB. These empirics now take up just 39 lines of text, plus Figures 3-5.

@@

We would like to have more mediators, but appropriate variables are hard to find. Earnings in first job reduces our N by a factor of 10, which makes results too noisy to be useful. As the reviewer suggests, current household income is very much endogenous not only to human capital, but also to previous fertility, which makes it inappropriate as a mediator.

> \> Also, in the mediation analysis, some of the ratios of indirect to total effects are negative or larger than 100%. This could indicate that the indirect and direct effects have opposite signs, but could also be due to the imprecision of the estimates. The authors should estimate 95% CIs for the ratio of indirect/total effects (these can be obtained by bootstrapping). They could then focus their discussion on the ratios that are significant and not on those that are just noisy estimates.

Bootstrapping ratios is not so simple, since if the confidence interval for the denominator contains 0, the ratio confidence set may contain infinity and may not be a single interval, and the standard approach of taking quantiles can be misleading (see Franz 2007). Alternatives like the Fieller method have the same problem. Another issue is that to estimate Bonferroni-corrected significance at p = 0.05/33 we'd need 0.075- and 99.925-percentiles of the distribution, which would require many thousands of bootstraps (infeasibly slow, given our sample size).

Following the advice of Franz (2007), we now compromise: we bootstrap uncorrected 95% CIs of the ratio, but only when the denominator (i.e. the total effect) is significantly different from 0 at $p$ \< 0.05/33. This is substantively sensible anyway: if we can't reject that a PGS has no effect on fertility, it does not make much sense to ask whether the effect is mediated. We continue to report p values for the indirect effect only, following Zhao et al. (2010).

@@

> \
> 3. To adjust for ascertainment bias, the authors adjust their estimates using three weighting schemes. Why not combine all the variables from the three schemes and add other observables that could impact ascertainment (including also sex, proximity to a recruitment center, number of individuals in the household, and household income)? If the authors' goal is to adjust their estimates for ascertainment in the UKB, they should use all relevant observables to create UKB weights so that their estimates more closely resemble those they'd obtain in a more representative sample. This is particularly important here since we're interested in selection, and estimates can be severely biased in a selected sample like the UKB.\
> \
> Also, instead of presenting a small subset of the results with various weighting schemes in Figure 2, it'd be preferable for the authors to pick what they think is the best weighting scheme and then present all results with that scheme. Then, they can briefly discuss the robustness of their results to various weighting schemes.

XXX Probably get in touch with those guys with their own weighting.\

> 4\. On p. 4, line 13, the authors write that "Beta is the expected polygenic score among children of the sample", with a footnote explaining the reason. However, since the authors use N=number of children rather than relative number of children (=number of children/mean number of children), I don't think that's correct. See Beauchamp PNAS 2016. More generally, why not use relative lifetime reproductive success (rLRS) instead of number of children as the dependent variable in equation 1?

Oops, yes. We hope the corrected footnote is now right.

We now use RLRS as the dependent variable for both respondents and parents. Results are not much changed, except there is now slightly more evidence for larger selection effects over time: 9 polygenic scores show a significant increase in effect size on respondents' RLRS when we median-split them by birth year. In turn, this makes us less sure that we can rule out that the welfare state is a relevant explanation for natural selection, so we have slightly qualified our comments on this topic.

@@

> 5\. How were the 33 PGSs selected? For instance, why didn't the authors include a PGS of risk preferences? (I'm not requesting that the authors include that specific PGS, but only that they explain their selection process.)

We address this in the Materials and Methods section: "Polygenic scores were chosen so as to cover a reasonably broad range of traits, and based on the availability of a large and powerful GWAS which did not include UK Biobank."

> 6\. Did the authors use an age cutoff to include males and females in their selection analyses. To ensure measured fertility is completed fertility, the authors should only use males who were at least 50 years old and females who were at least 45 years old at the time of measurement.

We used 45 years old. We have adjusted this to 50 male/45 female as suggested. Results are little changed, except that EA3 is no longer undergoing diversifying selection in the respondents' generation at the p = 0.05/33 significance threshold.

@@

> 7\. The authors write that sigma captures risk aversion in income in their model. But that's confusing and not a good way of describing this for their model, because as far as I understand there's no income risk! They should clarify this and say that sigma regulates the curvature of the utility function, and explain what this means in their riskless model.

It is a fair point. We have rewritten accordingly: "$\sigma > 0$ measures the curvature of the utility function, i.e. the decline in marginal utility of income as income increases".\
\
OTHER COMMENTS AND SUGGESTIONS

> 8\. Most figures are hard to read. For example, Figures 3-5 are hard to read. In Figure 3a, the dot color indicate education and the dot border color indicates significance; that's very difficult to read quickly (for instance, a blue dot with a yellow border looks green). Instead, the authors could use different symbols (crosses, squares) for education and colors for significance, or vice-versa.

We experimented a lot with these figures, and they may not be easy to improve. For example, we tried using symbols, but in our judgment the added visual noise makes it harder to see the pattern of the results across rows. Below we show three alternative versions of one of our figures: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes. We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor.

### 2022-03-14 why-natural-selection.Rmd
Shorten section 2, move some results to appendix.

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
          my_vline + 
          confint_geom_segment +
          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
          pval_scale() +
          my_fill_scale(n = n_cats, direction = fill_direction) +
          my_labs() +
          my_theme() +
          scale_linetype_manual(name = "", values = c("95% c.i." = 1))

@@

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i."), 
                    color = "grey60", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

@@

```{=tex}
\begin{equation}
RLRS_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

@@

[^normalization]: XXX The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
where $N$ is the number of children. Since PGS are normalized to variance 1 and
mean 0, this reduces to 
$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
score weighted by the number of children, which is the average polygenic score 
in the next generation [@robertson1966mathematical].

@@

We also check for stabilizing and disruptive selection by estimating
\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
we find disruptive selection for educational attainment polygenic scores: at
higher values of these scores, the negative effect on fertility is smaller
(Appendix Figure \@ref(fig:plot-purifying)).

[^balance-diverse-def]: Stabilizing selection selects for intermediate values, 
reducing the variance of the trait under selection. Disruptive selection
selects for extreme values and increases the variance.

```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score."}

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.5, stroke = 0.5) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
      my_theme()

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.
We emphasize that these categories are not necessarily exogenous to the polygenic
scores. For example, both in the data and in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related
polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
differences in equation \@ref(eq:regression) across subgroups constrain the set
of possible explanations for natural selection. As we show below, a simple
economic model predicts the differences we find.

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category.  These results could be
explained by age, if older respondents have lower income and are less educated,
and also show more natural selection on polygenic scores. However, when we rerun
the regressions, interacting the polygenic score with income category and also
with a quadratic in age, the interaction with income remains significant at
$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
if we interact the PGS with age of leaving full time education and a quadratic
in age, the interaction with age leaving full time education remains significant
at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix Figure
\@ref(fig:plot-sexes)). Differences are particularly large for educational
attainment, height and MDD. Several polygenic scores for mental illness and
personality traits are more selected for (or less against) among women,
including major depressive disorder (MDD), schizophrenia and neuroticism, while
extraversion is more selected for among men. Scores for waist circumference and
waist-hip ratio are less selected for among women. One
possible reason for these sex differences is that polygenic scores may affect
fertility via success in marriage markets, and men and women may value different
characteristics in these.

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).

@@

[^aflb-no-men]: This information is unavailable for men.

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}

@@

To investigate this further, we estimate equation \@ref(eq:regression) among
females, *controlling* for AFLB. In `r opp_signed` out of 
`r length(score_names)` cases, effects change sign when controls are added. The
correlation between effect sizes controlling for AFLB, and raw effect sizes, is
`r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: an effect on AFLB, and an opposite-signed effect on number of children
after AFLB is controlled for.

@@

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among either the parents' or
the respondents' generation. We also check whether selection effects vary by
socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between respondents' 
and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

Suppose that certain genetic variants correlate with human capital -- skills or
other characteristics that affect an individual's earnings in the labour market
[@mincer1958investment; @becker1964human]. These variants may then be associated
with opposing effects on fertility. The income effect will lead to natural
selection in favour of earnings-increasing variants (or variants that are merely
associated with higher earnings). The substitution effect will do the reverse.

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1, and let $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the curvature
of the utility function, i.e. the decline in marginal utility of income as income
increases. We examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with our corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of $\sigma$, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

@@

Thus, a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise.

------------------------------------------------------------------------------------
    **Theory**                                        **Empirical results**
--- ------------------------------------------------- ------------------------------
1.  The fertility-human capital                       Figures \@ref(fig:plot-means-over-time) 
    relationship is negative:                         and \@ref(fig:plot-res-wt).
    $\frac{dN^{*}}{dh} < 0$.

2.  The fertility-human capital                       Figure \@ref(fig:plot-income-educ-level)a. 
    relationship is weaker (closer to zero)           Selection effects are also weaker at higher
    at higher wages and/or levels of                  polygenic scores for 
    human capital.                                    educational attainment (Appendix Figure
                                                      \@ref(fig:plot-purifying)).

3.  The relationship is more negative when            Stronger effects for single parents
    the time burden of children, $b$, is              (Figure \@ref(fig:plot-n-partners)).
    larger.

4.  The relationship is weaker at higher              Figure \@ref(fig:plot-income-educ-level)b.
    levels of education $s$.

5.  The relationship is weaker among those            Effects weaker among those starting 
    who start fertility in period 2                   fertility later 
    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)).
    start fertility in period 1 
    ($N_{1}^{*} > 0$).
----------------------------------------------------------------------------------------

@@

```{r literature, eval = FALSE}
"Other models of the literature

## LITERATURE NOTES

* Lundberg and Pollak JEP 2007:
    - several models where nonmarital fertility is more likely for poorer women
    - Burdett and Ermisch 2002, Willis 1999, Neal 2004
    - Unmarried parents may not be able to credibly commit to each other

* Happel et al. 1984
   - childbirth is chosen to minimize loss of earnings from interrupted work.
   - early childbirth if your skills decay a lot  while childrearing. (No point
     gaining them then losing them.) Late childbirth otherwise.
   - late childbirth if you have acquired skills before marriage.
   - Capital-market imperfections make it wise to delay childbirth (till man has
     high income to cope while wife is at home). More so for men's wages which
     rise faster.
   - empirically: (among married couples) 
     - high-skill occupation women have children later. Higher-earning men have 
     them earlier.

- Caucutt et al. 2001
  - high waged and high educated women (of recent cohorts) more likely to
    delay childbirth

- Cohen et al. 2013
  - analyse effect of Israeli subsidy
  - along the way, find that income affects fertility negatively at low income,
    positively at high income, supporting Becker and Tomes 76

- Becker and Tomes 1976
  - investment in children quality and quantity when quality must be equal
    across children; and some quality comes from an innate endowment.
  - "observed" quantity-of-children income elasticity rises as income rises
  - while "observed" quality-of-children elasticity falls as income rises, 
    b/c more of total quality must be paid for by parents themselves. 
  - the need to pay more for quality results in lower quantity, since total cost is 
    quantity X quality
  - something fishy about argument somehow...

- Lindo 2008
  - men who lose a job have children sooner, but fewer children in total
  - controls for all skills effects
  - so overall effect of income is positive, but timing changes

- Monstad et al. 2008
  - using an education reform to vary education
  - no evidence of a causal effect on fertility, but births start later
  * maybe the usual correlation reflects underlying skills. Your data could
    test this.

- Autor et al. 2017
  - the China shock (exogenous shock to income) 
  - they use Chinese import penetration by industry (instrumented using
    other countries) times industry employment shares; and separate it for
    males and females
  - shocks to male employment decreased fertility and marriage; 
    increased unmarried mothers
  - shocks to female employment *increase* marriage and motherhood
  * this roughly seems to be the consensus
  * Whereas, you find that male and female *skills* (as 'instrumented' by PGS)
    have the same effect on fertility - i.e. less skills = higher fertility

- Baudin et al. 2015
  * look at childlessness vs. fertility (extensive/intensive margin)
  * some people are too poor to afford children
  * 'a husband alleviates the time cost of having children'. But for some groups (see appendix)
    there is a +ve income/fertility gradient
  * fertility decreases with female educn due to subst effect
  * baby boom and bust in US: a decline in childlessness caused by poverty, then an increase
    caused by substn effect ('opportunity cost')
  * fertility decreases with educn both for single mothers and married women
    - slightly (25%) higher decrease per year of educn (Table 2) for single mothers

@@

Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
analysis to directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
indirect effect of the score on fertility *via* educational attainment is
significantly different from zero ($p < 0.05/33$) and takes the same sign as the
overall effect. Among these scores, the median proportion of the total effect
explained by the indirect effect is `r percent(median_prop)`. Figure
\@ref(fig:plot-mediation) shows estimated proportions explained along
with bootstrap 95% confidence intervals (uncorrected; 100 bootstraps).
The educational attainment variable is a relatively crude measure of human 
capital: more accurate measures would likely explain more of the total effect.

@@

@@

@@

>>>>

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
          my_vline + 
          confint_geom_segment +
          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
          pval_scale() +
          my_fill_scale(n = n_cats, direction = fill_direction) +
          my_labs() +
          my_theme() +
          scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1))

@@

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                    color = "grey60", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

@@

```{=tex}
\begin{equation}
\mathrm{RLRS}_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

@@

[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$
where $N$ is the number of children. Since PGS are normalized to variance 1 and
mean 0, this reduces to 
$Cov(RLRS, PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the polygenic 
score weighted by relative lifetime reproductive success, which is the average polygenic score 
in the next generation [@robertson1966mathematical].

@@

```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score. P value threshold is 0.05, Bonferroni-corrected for multiple comparisons. Confidence intervals are uncorrected."}

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.5, stroke = 0.5) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. These results are robust 
to controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).[^sex]

[^sex]: We also compared selection effects between men and women (Appendix Figure
\@ref(fig:plot-sexes)).

@@

[^aflb-no-men]: AFLB is unavailable for men.

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4, fig.width = 4.5}

@@

To investigate this further, we estimate equation \@ref(eq:regression) among
females, *controlling* for AFLB. In `r opp_signed` out of 
`r length(score_names)` cases, effects change sign when controls are added. The
correlation between effect sizes controlling for AFLB, and raw effect sizes, is
`r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: a correlation with AFLB, and an opposite-signed correlation with 
number of children after AFLB is controlled for.

@@

We emphasize that our categories are not necessarily exogenous to the polygenic
scores. For example, in the data, and also in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related
polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
differences in equation \@ref(eq:regression) across subgroups constrain the set
of possible explanations for natural selection. As we show below, a simple
economic model predicts the differences we find.

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among the parents' generation.
There is some evidence that selection effect sizes are increasing in the 
respondents' generation, with 9 polygenic scores showing a significant increase. 
We also check whether selection effects vary by
socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between respondents' 
and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

Suppose that certain genetic variants correlate with *human capital*: skills or
other characteristics that affect an individual's earnings in the labour market
[@mincer1958investment; @becker1964human]. These variants may then be associated
with opposing effects on fertility. The income effect will lead to natural
selection in favour of earnings-increasing variants (or variants that are merely
associated with higher earnings). The substitution effect will do the reverse.

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1, and let $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the curvature
of the utility function, i.e. the decline in marginal utility of income as income
increases. We examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with our corresponding empirical results for the correlation
between polygenic scores and RLRS.[^one-period-results] The key insight of
the model is that for middling levels of $\sigma$, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

@@

Thus, a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise. There is also a literature showing negative correlations
between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].

------------------------------------------------------------------------------------
    **Theory: the fertility-human capital           **Empirical results**
    relationship is...**                                      
--- ----------------------------------------------- --------------------------------
1.  Negative: $\frac{dN^{*}}{dh} < 0$.              Figures \@ref(fig:plot-means-over-time) 
                                                    and \@ref(fig:plot-res-wt).

2.  Weaker (closer to zero) at higher wages and/or  Figure \@ref(fig:plot-income-educ-level)a.
    levels of human capital.                        Selection effects are also weaker at 
                                                    higher polygenic scores for 
                                                    educational attainment (Appendix Figure
                                                    \@ref(fig:plot-purifying)).

3.  More negative when the time burden              Stronger effects for single parents
    of children, $b$, is larger.                    (Figure \@ref(fig:plot-n-partners)).

4.  Weaker at higher levels of education ($s$).     Figure \@ref(fig:plot-income-educ-level)b.

5.  Weaker among those who start fertility          Effects weaker among those starting 
    in period 2 ($N_{1}^{*} = 0$) than among        fertility later 
    those who start fertility in period 1           (Figure \@ref(fig:plot-age-flb)).
    ($N_{1}^{*} > 0$).

----------------------------------------------------------------------------------------

@@

@@

Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
analysis to directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
the score on fertility *via* educational attainment takes the same sign as the
overall effect, and is significantly different from zero ($p < 0.05/33$). Among
these scores, the median proportion of the total effect explained by the
indirect effect is `r percent(median_prop)`. Figure \@ref(fig:plot-mediation)
shows estimated proportions explained along with bootstrap 95% confidence
intervals (uncorrected; 100 bootstraps). The educational attainment variable is
a relatively crude measure of human capital: more accurate measures would likely
explain more of the total effect.

@@

Code for this paper is available at 
https://github.com/hughjonesd/negative-selection.

@@

Figure \@ref(fig:plot-sexes) plots selection effects by sex. Differences are
particularly large for educational attainment, height and MDD. Several polygenic
scores for mental illness and personality traits are more selected for (or less
against) among women, including major depressive disorder (MDD), schizophrenia
and neuroticism, while extraversion is more selected for among men. Scores for
waist circumference and waist-hip ratio are less selected for among women. One
possible reason for these sex differences is that polygenic scores may affect
fertility via success in marriage markets, and men and women may value different
characteristics in these.

@@

\FloatBarrier
\clearpage
## Controlling for age {#sec:age-control}
